Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-06 3.Thesis/Writeup/2.Manuscript.Rmd
Emphasize unaffected univariate Qs

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s. **[Add panel labels, describe panels]** Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$. The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$. The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$. This implies that, under the current specifications, no more than seven iterations are necessary to obtain *sufficient* performance in terms of bias and confidence validity.

### $Q=\mu$.
The estimated univariate means seem unaffected by the number of iterations in the algorithm. This implies that *suffient* inferences may be obtained with as little as one iteration ($T \geq 1$). The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% causing more extreme bias than other conditions. Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$. Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is relatively small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$. **[Add: up-to 50% missingness everything is relatively unbiased, while 75% and 95% missingness show very different results. Why is that? Add sd of Y in the figure to show that the bias is relative: a bias of 0.02 for a variable with a mean of 25.81, and sd of 11.32 is nothing.]**

### $Q=\sigma$.
The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations. *Sufficient* inferences are obtained in conditions with just two iterations. Interestingly, the bias in first iteration is actually *less*  extreme than for any other number of iterations. We may therefore conclude that no iteration at all is needed to obtain *sufficient* estimates for $Q=\sigma$. Completely unbiased estimates are only obtained in conditions where $p_{mis}=.05$. Conditions where $p_{mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion. 
<!-- in the first iteration. how impact of non-convergence similar to the means. The magnitude of the bias $\bar{Q}$ follows the incremental order of the missingness proportions. The number of iterations in the algorithm has little effect on the bias. Only the estimates in simulation conditions with just one iteration ($T=1$) differ from other conditions that have the same missingness proportion but a higher number of iterations. Interestingly, the bias in conditions where $T=1$ is *less* severe than after continued iteration. For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{mis} = 0.05$ results in completely unbiased $\bar{Q}$s. -->

@@

We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest $Q$). The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence. We differentiate between approximately unbiased and confidence valid estimates, and *sufficiently* valid estimates. The former is defined by the interplay between $Q$, $p_{mis}$, and $T$. The latter is observed in all simulation conditions where $T\geq7$, irrespective of the $Q$ and $p_{mis}$. We thus expect the most signs of non-convergence in conditions where $T<7$.

@@

### $\theta=$ chain mean.
When $\widehat{R}$ and $AC$ are applied to the chain means, we notice no impact of the missingness proportions. The results only depend on the number of iterations per simulation condition.

>>>>

Figure \ref{fig:Qs} displays the influence of inducing non-convergence on the estimated quantities of scientific interest, $Q$s. **[Add panel labels, describe panels]** Overall, conditions with a higher proportion of missingness and/or a lower number of iterations are impacted more severely by non-convergence. I.e., on average these conditions portray more extreme bias in the estimated $Q$s, and non-nominal coverages. Roughly speaking, this means that the algorithm is indeed not converged at $t=1$, and converges gradually as $t$ increases to $T$. The point at which an additional iteration does not lead to an improvement in the estimates depends on quantity of scientific interest $Q$ and the missingness proportion $p_{mis}$. The highest number of iterations that is necessary to reach this point across all simulation conditions is $T=7$. This implies that, under the current specifications, no more than seven iterations are necessary to obtain *sufficient* performance in terms of bias and confidence validity. **[only substantial effect of non-conv in multivariate $Q$s]**

### $Q=\mu$.
The estimated univariate means seem unaffected by the number of iterations in the algorithm. This implies that *suffient* inferences may be obtained with as little as one iteration ($T \geq 1$). The magnitude of the bias in $\bar{Q}$ depends exclusively on the missingness proportion $p_{mis}$, with missingness proportions of 75% and 95% causing more extreme bias than other conditions. Unbiased estimates are obtained in all conditions where $p_{mis}\leq.50$. Note however, that even in the conditions where $p_{mis}\geq.75$, the bias is relatively small: $\mu_Y$ is maximally under-estimated by 0.02 units, while the true value of $\mu_Y$ is 25.81 ($\sigma_Y = 11.32$). We conclude that non-convergence does not substantially impact the validity of the inferences for $Q=\mu$.

### $Q=\sigma$.
The estimated standard deviations are largely affected by the missingness proportions, but not to much by the number of iterations. *Sufficient* inferences are obtained in conditions with just two iterations. Interestingly, the bias in first iteration is actually *less*  extreme than for any other number of iterations. We may therefore conclude that no iteration at all is needed to obtain *sufficient* estimates for $Q=\sigma$. Completely unbiased estimates are only obtained in conditions where $p_{mis}=.05$. Conditions where $p_{mis}\geq.25$ are impacted by non-convergence in order of increasing missingness proportion. Similar to the bias for $Q=\mu$ the magnitude of the bias in $Q=\sigma$ is not substantial in any condition. We conclude that both of the univariate $Q$s are practically unaffected by non-convergence.
<!-- in the first iteration. how impact of non-convergence similar to the means. The magnitude of the bias $\bar{Q}$ follows the incremental order of the missingness proportions. The number of iterations in the algorithm has little effect on the bias. Only the estimates in simulation conditions with just one iteration ($T=1$) differ from other conditions that have the same missingness proportion but a higher number of iterations. Interestingly, the bias in conditions where $T=1$ is *less* severe than after continued iteration. For the least possible bias, we may use any number of iterations $T\geq1$, but only $p_{mis} = 0.05$ results in completely unbiased $\bar{Q}$s. -->

@@

We evaluate the non-convergence diagnostics $\widehat{R}$ and $AC$ against the performance measures described above (i.e., bias and confidence validity of the quantities of scientific interest $Q$). The diagnostics should identify those simulation conditions that are negatively impacted by non-convergence. We expect the most signs of non-convergence in conditions where *sufficient* estimates are not obtained (high missingness proportions, $T<7$), and at least some signs of non-convergence in conditions where estimates are not approximately unbiased ($p_{mis}=.95$, any number of iterations).

@@

### $\theta=$ chain mean.
When $\widehat{R}$ and $AC$ are applied to the chain means, **we fail to identify the non-convergence in conditions where $p_{mis}=.95$**. We notice no impact of the missingness proportions. The results only depend on the number of iterations per simulation condition.
